Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-09-18 paint-contour-tree.tex
more changes to upper bound

We have a claim that also bounds heap sizes for special set of paths.

\begin{claim} \label{clm:above} Let $Q$ be a set of incomparable nodes in the shrub $P(T)$ (which are paths in $T$).
Suppose all paths in $A(Q)$ are heavy.
Then $\sum_{q \in Q} |H_q| \log |H_q| = O(\sum_{p \in A(Q)} |p|\log |p|)$.
\end{claim}

\begin{proof} Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.

Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
Each such $v$ must occur in some path $p \in A(Q)$. Thus,
$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
We apply the bound of \Lem{pathBounds} to complete the proof.
\end{proof}

@@

\begin{lemma} \label{lem:light} $\sum_{p \in L} \sum_{v \in p} \log |H_v| = O(\sum_{p \in L'} |H_p| \log |H_p|)$.
\end{lemma}

@@

\begin{proof} (of \Thm{runtime}) We split into heavy and light paths.
We have $\sum_{v \in T} \log |H_v| = \sum_{p \in L} \sum_{v \in p} \log |H_v| + \sum_{p \notin L} \sum_{v \in p} \log |H_v|$.
For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in L'} |H_p| \log |H_p|)$.
Now, we apply \Clm{above} to the set $L'$, noting that it satifies the claim conditions and that $A(Q) \cap L = \emptyset$.
Hence, the first term is $O(\sum_{p \notin L} |p|\log |p|)$.
The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
$\sum_{p \notin L} |p|\log |p| \leq \cost(\pmax(T))$.
\end{proof}

@@

\subsection{Upper Bound by Path Decomposition}

>>>>

There are some light paths that we can also ``pay" for.

\begin{definition} \label{def:support} A light path $p$ is called \emph{supported} if there exists
a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
and all $q_i$'s are heavy.
\end{definition}

Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths.
%We have a claim that also bounds heap sizes for special set of paths.

\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{p \in \cH} |p|\log |p|)$.
\end{claim}

\begin{proof} Pick $p \in \cL'$. As we traverse up $q_1, q_2, \ldots, q_k$, there
are certain vertices of $H_p$ that lie in these paths. There must be at least $|H_p|/100$
such vertices, since $H_{q_k} \leq 99H_p/100$. These are encountered in
a fixed order. Let $H'_p$ be the first half of these vertices. For all $v \in H'_p$
$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$
and $\sum_{p \in L'} |H_p|\log |H_p| = O(\sum_{p \in L'} \sum_{v \in H'_p} \log |H_v|)$.

Pick another $\widetilde{p} \in \cL'$. If $p$ and $\widetilde{p}$ are ancestor-descendant,
then their corresponding supporting ancestor chains are disjoint. Any $v \in H'_p$
can only occur in one other $H'_q$ (for $q \in \cL'$), because $v$ has exactly two colors.
In other words, any $v$ appears at most twice in the above double summation.
These vertices are all present in heavy paths, so we can bound
the double summation by $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
%
%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
%
%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
We apply the bound of \Lem{pathBounds} to complete the proof.
\end{proof}

@@

\begin{lemma} \label{lem:light} $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
\end{lemma}

@@

\begin{proof} (of \Thm{runtime}) We split into heavy and light paths.
We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
\end{proof}

@@

\subsection{Proving \Lem{light}}

We can simplify a little. We split the summation and apply \Lem{pathBounds}. Finally,
we note that for $p \in \cL$, $|p| \leq |H_p|$.
\begin{eqnarray*} 
\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \\
& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(|p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(|H_p|\log |H_p|)
\end{eqnarray*}
